### 2013-02-14 doc/xps2013/summary.tex

%% NSF XPS Proposal 2013

\input{header}

\begin{center}
  \textbf{\Large XPS-CLCCA: Scalable Parallelism for Irregular and Graph Applications}\\ \vspace{5mm}
  \textbf{Luis Ceze ~~~~~ Simon Kahan ~~~~~ Mark Oskin (PI)}\\
  \textbf{University of Washington}
\end{center}

\section{Introduction}

Irregular applications generate tasks with work, dependences, and memory
accesses that are highly sensitive to input. Classic examples of irregular
applications include branch and bound optimization, SPICE circuit simulation,
contact algorithms in car crash analysis, and network flow. Important
contemporary examples include processing large graphs in the business,
national security, machine learning, data-driven science, and social network
computing domains. For these emerging applications, fast response -- given the
sheer amount of data -- requires multinode systems. The most broadly available
multinode systems are those built from x86 commodity computing nodes
interconnected via ethernet or infiniband. Our focus is to enable scalable
performance of irregular applications on these mass market systems,
addressing two key challenges:

\paragraph{Irregular applications exhibit little spatial locality.} It is not
atypical for any given task's data references to be spread randomly across the
entire memory of the system. This makes current memory hierarchy features
ineffective. Caches are of little assistance with such low data re-use and
spatial locality. Commodity prefetching hardware is effective only when
addresses are known many cycles before the data is consumed or the accesses
follow a predictable pattern, neither of which occurs in irregular
applications. As a consequence, commodity microprocessors stall often when
executing irregular applications.

\paragraph{Irregular applications frequently request small amounts of off-node
data.} On multinode systems (e.g., clusters of computers), the challenges
presented by low locality are analogous, and exacerbated by the increased
latency of going off-node. Irregular applications also present a challenge to
commodity network technology, which is designed to transfer large blocks of
data, not the word-sized references emitted by irregular application tasks.
Injection rate into the network is insufficient to utilize wire bandwidth when
blocks are below about two-kilobytes, so any straightforward communication
strategy severely under-utilizes the network.

Luckily, many of the important irregular applications naturally offer large
amounts of concurrency. This immediately suggests taking advantage of
concurrency to tolerate the latency of data movement. The fully custom Tera
MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular
applications by using concurrency to hide latencies. It had a large
distributed shared memory with no caches. On every clock cycle, each processor
would execute a ready instruction chosen from one of its 128 hardware thread
contexts, a sufficient number to fully hide memory access latency. The network
was designed with a single word injection rate that matched the processor
clock frequency and sufficient bandwidth to sustain a reference from every
processor on every clock cycle. Unfortunately, while an excellent match to
extremely irregular applications, the MTA was not cost-effective on
applications that could exploit locality and had very poor single-thread
performance, making it a commercial failure. The Cray XMT approximates the
Tera MTA-2, substantially reducing its cost but not overcoming its narrow
range of applicability.

In this proposal we first describe our early
Grappa prototype, a runtime system for irregular
applications. It provides several facilities: a global shared memory
abstraction, user-level threading support, and communication and
synchronization primitives. Applications targeted for Grappa are not expected
to have \emph{any} locality (although what locality that exists is favorably
exploited), but are expected to provide many thousands of concurrent threads.

Next, we present some preliminary results. Even as an early prototype, Grappa
runs several graph-crunching applications (classic examples of irregular
behavior) efficiently on a commodity cluster. For unbalanced tree search,
Grappa is over 3X faster and shows greatly improved scalability compared to
the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower --
fast compared to implementations without Grappa, but obviously there is room
for improvement.

Given these very positive early results, this proposal plans to extend Grappa
in the following directions:

\begin{itemize}

\item \textbf{Reinvent network communication:} Our preliminary work has
revealed limitations of the current Grappa runtime system with regards to
network scaling. This proposal plans to enhance the network layer with whole
new forms of communication scheduling and aggregation, and explore new forms
of hardware/software interactions of remote direct memory access (RDMA).

\item \textbf{Language and compiler support:} Presently, Grappa is implemented
strictly as a library. Programmers must express remote operations explicitly
as function calls. This adds burden high-level language programmers do not
usually face and obfuscates opportunities for compiler optimizations. A
significant research thrust of the proposed work is to use language and
compilation support to provide a cleaner programming model and use
statically-available knowledge to lessen the runtime overhead automatically.

\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a
comprehensive benchmark suite of irregular and graph applications. We
are also seeking funding through this proposal to put together just
such a benchmark suite. This suite will include benchmarks, reference
implementations, and input data-sets. In addition a website will be
created to distribute freely the benchmark suite and publish results.

\item \textbf{Support for \ssd\ only data structures:} The size of data being
processed by large systems today (hundreds of terabytes) dwarfs what can fit
into a single system. Multiple system nodes are required to maintain and
access all of this data efficiently. However, given the lack of locality in the application, loading this data into DRAM is wasteful, in addition to adding overheads. The
Grappa runtime is designed to make applications tolerate extremely high memory
latency, thus we will explore having program data structures reside in solid state disks only. This further takes advantage of Grappa's latency tolerance properties to deal with larger data sets and reduce system cost.

\end{itemize}

%%%%%%%%%%%%%% Tempting to delete from here....

Finally, we propose to explore how the lessons learned performing this research -- specifically knowledge of what ultimately limits Grappa's performance for irregular applications using x86 processors and commodity networks with modified communication protocols -- can be applied to the design of future architectures.  Doing so will include studying the following research questions:

\begin{itemize}
\item \textbf{What is the right system architecture support for irregular applications, including architecture, communication and storage?}
\item \textbf{What are the necessary language and compiler support for large-scale, data-intensive irregular applications?}
\item \textbf{How do we comprehensively evaluate performance of graph analytics systems?}
\end{itemize}

%%%%%%%%%%%%%%%%%  To here.... -Mark

@@

Efficiently executing irregular computations on commodity hardware is a
\emph{difficult} problem. Much about commodity hardware is seemingly designed
for the opposite application characteristics: caches assume applications have
locality; networks are built with low injection rates but high bandwidth,
assuming large packet sizes; even runtimes and operating systems are built
assuming relatively infrequent communication and synchronization compared to
computation. This proposal rests on the seemingly outlandish claim that this
commodity hardware / software stack can be coerced, through another layer of
software, into efficiently executing irregular applications that both lack
locality and communicate frequently with small amounts of data. A folksy way
of saying it is we think we can jam a square peg through a round hole.
Fortunately we have some preliminary evidence we won't get wedged along the
way. This is an intellectually challenging endeavor, that we believe is well
suited for the NSF XPS program. In summary, our goal is to exploit parallelism
from irregular computations to efficiently execute them at scale.

@@

Irregular computations are of increasing economic importance.  Moreover, they are key to several national security applications.  If commodity hardware can be cajoled into executing such applications efficiently it will open the door for new and rich applications.  Furthermore, irregular applications will then reap the benefits of advances in commodity processors and networks, rather than requiring custom hardware that advances only at much higher cost.  Graph queries, machine-learning and data-driven science will significantly benefit from the technology we aim to create from the ideas in this grant proposal.  In addition our work will include the construction of a comprehensive irregular benchmark suite, something to help guide our own and the field's research.

%% Roadmap

\section{Description of the Current Grappa System}

In the twenty years that have elapsed since the Tera MTA, commodity microprocessors have become much faster and multicore has driven down the absolute price of computation; commodity network price-performance has improved as well. This shift has afforded us the opportunity to attack the challenges posed by irregular applications by emulating in software and inexpensive mass market hardware, the approach taken by Tera. We exploit the increased aggregate instruction rate per socket relative to chip bandwidth, using what would otherwise be wasted instructions to manage the multiplexing of as many as several thousand tasks per core, thus tolerating memory latency, reducing stalls, and making better use of available bandwidth. Ultimately, the opportunity is to cover the spectrum of irregular to regular computation: where tasks exhibit locality, multiplex fewer tasks and expend fewer instructions on context switching; where locality is lacking, multiplex more tasks at a higher rate to tolerate latency. Thus we make the best use of task parallelism -- either to scale to more cores or to tolerate latency -- and of caches -- either to exploit application locality or to house more task contexts.

Grappa is a software runtime system that allows a commodity x86 distributed-memory HPC cluster to be programmed as if it were a single large shared-memory machine and provides scalable performance for irregular applications. Grappa is designed to smooth over some of the performance discontinuities in commodity hardware, giving good performance when there is little locality to be exploited while allowing the programmer to exploit it when it is available.

Grappa leverages as much freely available and commodity infrastructure as possible. We use unmodified Linux for the operating system and an off-the-shelf user-mode infiniband device driver stack~\cite{OFED}. MPI is used for process setup and tear down. GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations. To this commodity hardware and software mix Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as few as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} and synchronizing operations including \emph{fetch-and-add\/}~\cite{fetchandadd} as well as user-defined atomic methods; and (3) a \emph{message aggregation\/} layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time. As we will show later, Grappa can tolerate latencies way beyond that of the network. Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our effective random access memory bandwidth, our synchronization bandwidth, and our ability to improve load imbalance.

\begin{figure}[t]
\begin{center}     
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth} 
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

The current Grappa prototype (Figure~\ref{fig:grappa}) has three main software components: \begin{description}

\item [Tasking system.] Our tasking system supports lightweight multithreading to tolerate communication latency and global distributed workstealing (i.e., tasks can be stolen from any node in the system), which provides automated load balancing.

\item[Distributed shared memory.] Our DSM system provides support for fine-grain access to data anywhere in the system. It supports synchronization operations on global data, explicit local caching of any memory in the system, and support for operation on remote data (delegating operations to home node). By tight integration with the tasking system and the communication layer, our DSM system offers high aggregate random access bandwidth for accessing remote data.

\item[Communication layer.] As discussed earlier, modern commodity networks support high bandwidth only for large messages. Since irregular applications tend to need frequent communication of small requests, the main goal of our communication layer is to aggregate small messages into large ones to better exploit what the network can offer. It is largely invisible to the application programmer.

\end{description}

\subsection{Tasks}

The basic unit of execution in Grappa is a {\em task}. Each task is represented by a function pointer and its arguments. A new task is enqueued at spawn time; when resources are free, it is allocated a stack, bound to a core, and executed.

During execution, a task yields control of its core following initiation of any long-latency operation, allowing the processor to remain busy while waiting for the operation to complete.  In addition,  a programmer can direct scheduling explicitly via the Grappa API calls. To minimize yield overhead, the Grappa scheduler operates entirely in user-space and does little more than store register state of one task and load that of another. Context switch times are as low as 38ns even when switching amongst thousands of tasks.

\subsection{Expressing parallelism}

Grappa programmers focus on expressing as much parallelism as possible without concern for where it will execute. Grappa then chooses where and when to exploit this parallelism, scheduling as much work as is necessary on each core to keep it busy in the presence of system latencies and task dependences.

Grappa provides two methods for expressing parallelism. First, when the programmer identifies work that can be done in parallel, the work may be wrapped up in a function and queued with its arguments for later execution using a \texttt{spawn}. A programmer can further specify using \texttt{spawn\_on} to spawn a task on a specific core in the system or at the home core of a particular memory location. Second, the programmer can invoke a parallel for loop, provided that the trip count is known at loop entry. The programmer specifies a function pointer along with start and end indices and an optional threshold to control parallel overhead. Grappa does {\em recursive decomposition} of iterations, similar to Cilk's cilk\_for construct~\cite {cilkforimplementation} and TBB's {\tt parallel\_for}~\cite{intel_tbb}.  It generates a logarithmically-deep tree of tasks, stopping to execute the loop body when the number of iterations is below a specified threshold.

\subsection{Memory}

Applications written for Grappa utilize two forms of memory: local and global.  Tasks executing on the same core share an address space local to that core.  The core accesses this local memory through conventional pointers, such as those generated by a standard C compiler.  We nonetheless refer to these pointers as ``local pointers'' to differentiate them from ``global pointers'' that may be used by other cores to reference the same physical memory as described below.  Applications use local memory for a number of things in Grappa: the stack associated with a task, localized global memory in caches (see below), and accesses to debugging infrastructure that is local to each system node.  Local pointers cannot access memory on other cores, and are valid only on their home core.

Large data that is expected to be shared and accessed with low locality is stored in Grappa's global memory. All global data must be accessed through Grappa's API. %, shown in Figure~\ref{fig:accessing-memory}.

\paragraph{Global memory addressing} Grappa provides two methods for storing data in the global memory. The first is a distributed heap striped across all the machines in the system in a block cyclic fashion.  Addresses to memory in the global heap use \textbf{linear addresses}.  Choosing the block size involves trading off sequential bandwidth against aggregate random access bandwidth. Smaller block sizes help spread data across all the memory controllers in the cluster, but larger block sizes allow the locality-optimized memory controllers to provide increased sequential bandwidth. The block size, which is configurable, is typically set to 64 bytes, or the size of a single hardware cache line, in order to exploit spatial locality when available. The heap metadata is stored on a single node. Currently all heap operations serialize through this node; while this has been sufficient for our benchmarks, in the future Grappa will provide parallel performance through combining~\cite{MAMA,flatcombining}.

Grappa also allows any local data on a core's stacks or heap to be exported to the global address space to be made accessible to other cores across the system. Addresses to global memory allocated in this way use \textbf{2D global addresses}.  This uses a traditional PGAS addressing model, where each address is a tuple of a rank in the job (or global process ID) and an address in that process. The lower 48 bits of the address hold a virtual address in the process. The top bit is set to indicates that the reference is a 2D address (as opposed to linear address). This leaves 15 bits for network endpoint ID, which limits our scalability to $2^{15}$ endpoints. Any node-local data can be made accessible by other nodes in the system by wrapping the address and node ID into a 2D global address. This address can then be accessed with a delegate.  At the destination the address is converted into a canonical x86 address by replacing the upper bits with the sign-extended upper bit of the virtual address. 2D addresses may refer to memory allocated from a single processes' heap or from a task's stack.

\paragraph{Global memory access} There are two general approaches Grappa applications use to {\emph access} global memory. When the programmer expects a computation on shared data to have spatial locality to exploit, explicit {\em cache} operations may be used. When there is no locality to exploit, remote {\em delegate} operations are used.

\textbf{Explicit caching.} Grappa provides an API to fetch a global pointer of any length and return a local pointer to a cached copy of the global memory.  Grappa cache operations have the usual read-only and read-write variants, along with a write-only variant used to initialize data structures. Languages for distributed shared memory systems have done optimizations to achieve a similar goal. For example, the UPC compiler coalesces struct and array accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's message vectorization hoists small messages out of a loop \cite{FortranD:1992}.

Caching in Grappa additionally provides a mechanism for exploiting temporal locality by operating on the data locally --- even though we aim Grappa at applications with low locality, we want to be able to exploit it when available. 
Currently, the burden is on the programmer to maintain coherency by introducing explicit synchronization.  Future work will develop a software directory based coherence scheme to simplify consistent access to global data.

\textbf{Delegate operations.} When the access pattern has low-locality, it is more efficient to modify the data on its home core rather than bringing a copy to the requesting core and returning it after modification. Delegate operations provide this capability. Applications can dispatch computation to be performed on individual machine-word sized chunks of global memory to the memory system itself (e.g., \emph{fetch-and-add}).  Delegate operations, proposed in \cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the primary synchronization method in Grappa.

\subsection{Communication}
\label{sec:communication}

In order to mitigate the low message injection rate limits of commodity networks, Grappa's communication stack has two layers: one for user-level messages and one for network-level messages.

At the upper layer, Grappa implements asynchronous active messages \cite{vonEicken92}. Each message consists of a function pointer, an optional argument payload, and an optional data payload. When a task sends a message, the message is copied to a send queue associated with the message's destination and the task continues execution.

Grappa's lower networking layer aggregates the upper layer's messages to
improve performance. Commodity networks including
Infiniband~\cite{infinibandbandwidth} achieve their peak bisection bandwidth
\emph{only} when the packet sizes are relatively large---on the order of
multiple kilobytes. The reason for this discrepancy is the combination of
overheads associated with handling each packet (in terms of bytes that form
the actual packet, processing time at the card and processing on the CPU
within the driver stack). Our measurements confirm manufacturers published
data~\cite{infinibandbandwidth}, that with this packet size the bisection
bandwidth is only a small fraction, less than 3\% of the peak bisection
bandwidth.

In our early experiments the vast majority of requests were smaller than 44
bytes, far too small to make efficient use of the network. To make the best
use of the network, we must convert our small messages into large ones. When a
task sends a message, it is not immediately sent, but rather placed in a queue
specific to the destination. If the size in bytes of a queue is above a
threshold (e.g., 4096, typically large enough for good bandwidth), the
contents of the queue are sent immediately. Also, each queue has a wait time
threshold ($\approx${1ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent immediately,
even if the queue size is lower than the message size threshold. Finally,
queues may be explicitly flushed, for situations when latency is more
important than bandwidth.

Underneath the aggregation layer, Grappa uses the \gasnet~communication
library~\cite{gasnet} to actually move data. All interprocess communication,
whether on or off a cluster node, is handled by the \gasnet~library.
\gasnet~is able to take advantage of many communication mechanisms, including
ethernet and infiniband between nodes, as well as shared memory within a node.

Some networks provide access to a remote machine's memory directly. This would seem to be a good fit for a programming model focused on global shared memory, however, we do not currently use it. In our experiments, we found that RDMA operations are subject to the same message rate limitations as all other messages on these cards, and thus using raw RDMA operations does not measurably increase achieved bandwidth (i.e., we are limited by injection rate not peak bandwidth).  At the moment, we implement remote memory operations with active messages in conjunction with aggregation, providing an order of magnitude or more increase in throughput compared to straightforward RDMA. As we describe later, however, we wish to explore richer semantics for RDMA communication that will make it integrate better with user level runtime systems such as Grappa.

\subsection{Performance}

To evaluate the performance of our current prototype with respect to the XMT,
we ran each of our three benchmarks on up to 16 nodes of each machine. Grappa
used 6 cores per node, with the best parameters chosen for each point. In some
cases, the XMT could not run the benchmark with 2 nodes, so the point is
omitted.

\begin{figure}
\begin{center}

\hspace{-1in}\begin{minipage}{0.3\textwidth}
\begin{center}
\includegraphics[width=2.5in]{figs/uts_performance.pdf}
\caption{\label{fig:uts_compare} Performance of in-memory unbalanced tree search.}
\end{center}
\end{minipage}
\hspace{1.25in}\
\begin{minipage}{0.3\textwidth}
\begin{center}
\includegraphics[width=2.5in]{figs/bfs_performance}
\caption{\label{fig:bfs-performance} Performance of breadth-first search.}
\end{center}
\end{minipage}

\end{center}
\end{figure}

\paragraph{Unbalanced tree search} We ran UTS-mem with a geometric 100M-vertex
tree (T1L). Figure~\ref{fig:uts_compare} shows the performance in terms of
number of vertices visited per second versus number of compute nodes. Grappa
is 3.2 times faster than the XMT at 16 nodes. As we will show later, the
performance advantage Grappa has over XMT increases as more nodes are added.
The main reason Grappa performs better is the software-based delegate
synchronization obviates the need for the retry-based synchronization that XMT
uses.

\paragraph{BFS} We ran BFS on a synthetic Kronecker graph with $2^{25}$ vertices and $2^{29}$ edges (25 GB of data). Figure~\ref{fig:bfs-performance} shows our performance in terms of graph edges traversed per second. The XMT is 2.5 times faster than Grappa at 16 nodes.  Performance has not peaked for Grappa, suggesting that adding more nodes will increase performance.

\section{Related work}

Our efforts are built on many existing ideas in programming languages, systems and architecture. In this section we discuss related frameworks and key enabling technologies that Grappa builds upon.

\paragraph{Comparable frameworks} Distributed graph processing frameworks like Pregel~\cite{pregel:2010} and Distributed GraphLab~\cite{distgraphlab:vldb12} share similar goals as Grappa. Pregel adopts a bulk-synchronous parallel (BSP) execution model, which makes it inefficient on workloads that could prioritize vertices. GraphLab, on the other hand, schedules vertex computations individually, allowing prioritization, which gives faster convergence in a variety of iterative algorithms.  GraphLab, however, imposes a rigid computation model where programmers must express computation as transformations on a vertex and its edge list only, with information only from adjacent vertexes. Pregel is only slightly less restrictive, as the input data can be any vertex in the graph.  Grappa also supports dynamic parallelism with asynchronous execution, but parallelism is expressed as tasks or loop iterations, which is a far more general programming model for irregular computation tasks.

\paragraph{Global memory} Grappa includes a custom implementation of a software distributed shared memory (DSM) system. Many traditional software DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact that they are built in software from applications by exploiting the processor's paging mechanisms, therefore relying heavily on locality. Instead, Grappa, like other partitioned global address space (PGAS) models, implements its DSM at the language, rather than system level.
%%note:  Myers suggestion is not incorporated here, as this seems ok:
Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005} make accesses to shared structures look like normal memory references. As we describe later, Grappa chooses a middle ground, where global addresses are explicit in the API and local accesses are emitted conventionally by the compiler.  While Grappa's DSM system is conceptually similar to prior work, its implementation is tuned for irregular computations.  Past DSM work, being page-based, could exploit the RDMA capabilities of network hardware to move large page-sized blocks of data from node to node. In our experience, when these networks move small blocks of data (a few bytes), only a fraction of the available bandwidth is achieved.
%%note:  Myers suggestion is not incorporated here, as this seems ok:
In addition, the DSM system in Grappa ends up being tightly coupled to the task scheduler in order to overlap long latency memory operations with useful computation.  For these two reasons it became necessary to build a new DSM system specifically for Grappa.

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory latency. This is a well known technique. Hardware implementations include the Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife}, Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in this paper, Grappa use a lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks on a single processing core. The large number of tasks is required because of the extremely high internode latency Grappa mitigates.  Grappa's task library employs several optimizations: an extremely fast task switch, a small task size, and judicious use of hardware prefetching to bring task state into the cache before the task is scheduled to execute.

\section{Proposed Research}

As described earlier, our work on Grappa so far has focused on building an initial prototype so we have a baseline system to build upon. This early prototype has lead us to think of many research opportunities, from new ways to improve network communication scheduling, to new data storage management ideas to compiler and language support. Below we detail our research plans in each of these thrusts.

\subsection{Networking}

Since Grappa is designed for applications that have no or low locality, nearly \emph{every} memory reference is to a remote system node.  This makes the networking layer key to overall system performance.  In this section we outline the key directions we wish to take in enhancing the networking layer of the Grappa runtime system.

\paragraph{Virtual topology:} Currently Grappa uses a flat network topology where each processing core in the system is individually addressed. As more nodes are added to the system, a quadratic increase in the number of threads (in aggregate across all nodes) is required in order to compensate for the \emph{same} network delay.  This is because the current aggregator treats each of these end-points individually, i.e., it only aggregates messages that go to the same exact end-points. This does not scale as the number of nodes increases because the probability of messages going to the same destination goes down (assuming a somewhat uniform communication distribution). In summary, this past design choice is simple but does not scale.

We need to rewrite the aggregator to build in a hierarchical virtual network layer.  The first layer of this hierarchy is to treat all processors at a given system node as a single destination. We intend to continue to use the lock-free single-threaded \gasnet~communication layer that we have been using, as we have found the lock-based one to incur too high of an overhead for our purposes.  Thus the aggregator will need to very carefully utilize lock-free data-structures and interact with \gasnet~carefully.
%%Didn't understand Myers suggestion here.

The second, and upper layers of the virtual topology are built from clustering system nodes into groups.  Instead of aggregating for all end-points, aggregation will occur for system nodes within a group.  A designated (and dedicated) node will act as a router, processing incoming and outgoing traffic to system nodes within the group and other groups.

We expect this hierarchical organization will reduce scaling requirements from
\BigO{n^2} threads to \BigO{n \times log(n)}. Of course, it takes what used to
be a single network hop for communication and turns it into \BigO{log(n)}
hops. More threads will be required to overcome this increased latency, but we
hypothesize only \BigO{log(n)} more, hence a real \BigO{n / log(n)^2} scaling
benefit is achievable with a virtual hierarchical network topology.

\paragraph{Hardware support for aggregation:} If the increased latency caused
by extra hops in hierarchical aggregation turns out to be detrimental, we
will explore mitigation by moving some of the aggregation logic to the network
card, either by exploiting off-the-shelf programmable network cards or
experimental systems such ask NetFPGA~\cite{netfpga} (while we are planning to
focus on Infiniband-based networking, if we explore the NetFPGA route we will
be using Ethernet).

\paragraph{User-level RDMA:} Certain commodity networking hardware today
supports remote direct memory access (RDMA). Unfortunately, \emph{how} it is
implemented is so primitive and low-level that it is difficult or impossible
to efficiently use in practice. We start with a short description of the
current semantics of RDMA implementations. Next we describe the research we
wish to undertake to make RDMA more useful.

Networking infrastructure supports remote read and write operations.  Read operations take a source address and key (on a remote node), a destination address and key (on the local node), and a length.  Cards support address translation for regions of memory that are registered with them.  These translations, indexed by the supplied key, are used to perform virtual to physical address translation.  The physical address of the destination address is similarly translated using the translation table at the local node.  Cards also support remote write operations with a similar set of semantics, where addresses are translated on a card using the card's local translation table.  Both read and write operations are asynchronous.  The requesting processor initiates the operation and then can check for completion by querying a completion queue at the local card.  \emph{Notably}, the processor(s) on the remote node are provided a polling interface whereby they can learn if a remote write operation occurred.  No such interface exists to learn about remote read or atomic operations.

This interface suffers from multiple shortcomings.  The biggest of which is that ultimately, moving data to and from a remote node constructively requires higher level semantic operations than simple reads and writes.  These higher level semantics can, of course, be synthesized from the underlying read and write primitives, but it wastes round trips and synchronization time.  One possible alternative solution is to enable remote nodes to dispatch high-level data-structure manipulation requests (e.g. insert this item into a queue, remove an item from a set).  The challenge here is to make those calls programmable and interoperate securely with a modern operating system, and a user-level runtime system.  Additionally there are questions about \emph{who} should carry out the computation.  Classic RDMA does not interrupt the host processor, but these richer programmable semantics will require either a processor in the network interface (necessitating issues such as coherence be managed) or complex RDMA tasks need a mechanism to interrupt the host processor \emph{without leaving user space}.

%% The second half of this paragraph is really confusing, I have just a minimal clue of what it is talking about. And the little I understand sounds a bit vague and not directly related to grappa.
The second challenge is that the virtual to physical address translation on the cards requires the operating system to pin translations.  This removes paging flexibility from the operating system, because runtime systems are (rationally) built to expect data structures to be in memory capable of RDMA or not.  Those in RDMA memory must be pinned.  In the long run, the solution here likely lies at the intersection of what progress is occurring in processor support for virtual machines and virtual memory translation for I/O.  Our goal will be to study carefully the I/O virtualization proposals put forth by major processor manufactures and devise solutions that enable RDMA to capitalize on these developments.

\subsection{SSD-only data structures}

Grappa applications are likely to use large amounts of data, 100s of terabytes to petabytes. Keeping all these data in DRAM only is unlikely to be cost effective. We will explore mechanisms that use solid-state storage as primary memory, greatly expanding the directly addressable storage at low cost.

Grappa is very effective in hiding latencies of network communication.  If an access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If the access is to memory stored on another node, Grappa issues a message to that node to request the data and context switches to other work until the reply arrives.  In fact, our measurements and estimates show that we might be able to tolerate even greater latency -- on the order of milliseconds per request.  This provides the tantalizing possibility that we will be able to keep data in \ssd\ drives only and service data access requests directly from them, without leaving copies in DRAM.  The natural point of integration for non-volatile memories in Grappa is to store the global distributed shared heap in non-volatile memory. All accesses to this shared heap are done through an \api, so this switch would be transparent to the programmer.

This is not straightforward, however, for three reasons: (1) \ssds\ assume spatial locality. In our benchmarks, we find the average request size to be less than 64 bytes. Today's \ssds\ are optimized for reads in the multi-kilobyte range. (2) \ssds\ have a low request rate. The fastest~\cite{fusionio} can perform on the order of 1 million random reads per second, and many are limited to less than 100K/s. Even the underlying flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, our measurements suggest cores may be idle in Grappa if each node cannot perform 100 million random reads per second of the global heap. And (3) \ssds\ have high overhead. Mass market \ssds\ are usually treated as normal disks: accesses must occur through the kernel, incurring a high context-switch overhead. Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.

To address this challenge we will explore building a flash array consisting of several ten's to a hundred disks \emph{per node}.  This will enable us to increase random accesses to our target 100 million references per second. But it will require several active disk controllers per system.  We do not expect any existing device driver stack will work at this scale.  Hence we will need to write a custom disk controller device driver and file system that supports such a high degree of concurrent requests across so many parallel channels.

%% Sorry Mark, while the item below is clearly rich and juicy, we can't leave it bhere because it is huuuuge, it does not match the scope of the grant, and we dont have the expertive --- e.g., are you sure nobody is working on it?

%% I think this is the only idea that is going to work... maybe we should delete the above three and make this one good.  Seriously, the ideas up there are bunk.  -ark
% \paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.  We will explore new flash microarchitectures that optimize for message rate as well as bandwidth.  Such chips have value far beyond Grappa.  Research into persistent key/value stores, and really any persistent high-throughput data-structure will benefit from a redesigned flash chip that can sustain dramatically increased message rates.

\subsection{Compiler and language support}

Language and compiler support are instrumental in making Grappa usable to a
wide audience of application developers. We plan to use the LLVM~\cite{llvm}
infrastructure. Below we describe the specific aspects we plan to tackle.

\paragraph{Relaxed consistency:} Presently Grappa uses a consistency model that is sequentially consistent for data-race free programs.  Underlying hooks are available to build even more relaxed consistency models, although we don't expect programmers themselves will want to use them.  A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency and hiding more latency.

\paragraph{Language support for synchronization operations: } At the moment, runtime library support is provided in Grappa for basic synchronization primitives (lock, barrier, etc).  Building synchronization directly into the language has the advantage that the compiler can understand and optimize it.  Several optimizations are possible in the Grappa framework, including exploiting the scheduler and work stealer to elide certain lock operations (scheduling is co-operative on a per-CPU basis), converting synchronization primitives into remote delegate operations, and exploiting the check-in / check-out nature of global memory in Grappa to coalesce multiple synchronization operations together.

\paragraph{Delayed writes: } At the moment when we we write applications for Grappa we use a technique that we call feed-forward delegates.  Essentially what this technique does is if a loop is reading and writing disjoint sets of data, it is possible and we have found efficient, to dispatch writes asynchronously and ensure those writes are completed only by the end of the loop.  This increases performance by further overlapping computation with memory access.  A compiler aided by programmer annotations or capable of alias analysis can do this transformation automatically for the programmer.

\paragraph{Automatic delegate synthesis: } Delegates are a technique in Grappa where one thread can request a small computation to be carried out on a remote node.  For instance, an increment on a memory location is dispatched to the system node that manages that memory operation and carried out there.  Currently the decision of what is a delegate is left to the programmer.  But this is an area rich for compiler optimization.  We will explore techniques to automatically synthesize delegates from code, potentially leading to more complex delegates and more efficient code.

>>>>

\paragraph{Summary}

Irregular applications are increasing in importance to the computing industry. For a long time they have underpinned computations of interest to the national security arm of the federal government. But now, they underpin such questions as ad placement in social networks, and analysis of complex data-sets in medicine and science. The defining characteristic of these applications is poor locality and massive available parallelism. Our effort is focused on making these applications perform well and scale on commodity hardware. The key idea that makes this work is to rely on massive concurrency to tolerate memory latency, instead of relying on locality (which doesn't exist for these applications). This core idea underpinned the XMT system, a fully-custom hardware system. We plan to explore that idea in software only, running on commodity processors. And unlike the XMT, however, we must also contend with commodity networking hardware, and here we again rely on threading to provide a sufficient number of requests that we can buffer them prior to dispatching on the network, to overcome bandwidth limitations when operating with small messages.

Our preliminary effort has shown a workable proof of concept of this approach: a commodity cluster can out do the XMT at its own game, by emulating its key features in software. What is needed now is additional effort to improve the underlying performance of this runtime and make it usable to a broad base of users. We will be focusing our efforts in the next three years on the key performance component, the networking layer, and the key programmer efficiency linchpin, high level language support. In addition we need a useful mechanism for handling data-sets in the petabyte range, and here we imagine building a cluster of machines that relies primarily on SSD's and not DRAM as main memory data-set storage.

@@

Efficiently executing irregular computations on commodity hardware is challenging.  Much about commodity hardware is seemingly designed for the opposite application characteristics: caches assume applications have locality; networks are built with low injection rates but high bandwidth, assuming large packet sizes; even runtimes and operating systems are built assuming relatively infrequent communication and synchronization compared to computation.  This proposal rests on the seemingly outlandish claim that this commodity hardware / software stack can be coerced, through another layer of software, into efficiently executing irregular applications that both lack locality and communicate frequently with small amounts of data.  A folksy way of saying it is we think we can jam a square peg through a round hole.  Fortunately we have some preliminary evidence we won't get wedged along the way.  This is an intellectually challenging endeavor, that we believe is well suited for the NSF XPS program.  In summary, our goal is to exploit parallelism from irregular computations to efficiently execute them at scale.

@@

Irregular computations are of increasing economic importance.  Moreover, they are key to several national security applications.  If commodity hardware can be cajoled into executing such applications efficiently it will open the door for new and rich applications.  Furthermore, irregular applications will then reap the benefits of advances in commodity processors and networks, rather than requiring custom hardware that advances only at much higher cost.  Graph queries, machine-learning and data-driven science will be enabled by the technology we aim to create from the ideas in this grant proposal.  Thus a broad impact of our work will be the transformative effect on the types of scientific and computing questions that commodity hardware can efficiently answer. In addition our work will include the construction of a comprehensive irregular benchmark suite to help guide our own and the field's research.
